---
title: "Assignment 1"
format: html
---

## 3. Exploratory Data Analysis with TEDS2016

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(haven)
library(janitor)
library(skimr)
library(MASS)

TEDS_2016 <- read_dta(
  "https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true"
)
```

```{r}
glimpse(TEDS_2016)
skim(TEDS_2016)
```

The initial EDA inspects variable structure, data types, and missingness patterns before modeling.

## 4. Problems encountered when working with the dataset

```{r}
# Check candidate variables used in this assignment
vars_needed <- c(
  "Tondu", "female", "DPP", "age", "income", "edu", "Taiwanese", "Econ_worse", "votetsai"
)

intersect(vars_needed, names(TEDS_2016))
setdiff(vars_needed, names(TEDS_2016))
```

## 5. How to deal with missing values

```{r}
# Convert haven-labelled vectors to plain vectors first
TEDS_clean <- TEDS_2016 |>
  mutate(across(where(haven::is.labelled), haven::zap_labels))

recode_missing_codes <- function(x) {
  if (!is.numeric(x)) return(x)
  x |>
    na_if(7) |>
    na_if(8) |>
    na_if(9) |>
    na_if(97) |>
    na_if(98) |>
    na_if(99)
}

TEDS_clean <- TEDS_clean |>
  mutate(across(everything(), recode_missing_codes))
```

```{r}
# Missing values in assignment variables
TEDS_clean |>
  summarise(across(all_of(intersect(vars_needed, names(TEDS_clean))), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

In this assignment, I handle missing values by putting known nonresponse codes to `NA`, then using `drop_na()` only on variables required by each specific model/table.

## 6. Relationship between `Tondu` and other variables

### Prepare analysis data and assign labels

```{r}
tondu_labels <- c(
  "Unification now",
  "Status quo, unif. in future",
  "Status quo, decide later",
  "Status quo forever",
  "Status quo, indep. in future",
  "Independence now",
  "No response"
)

analysis_vars <- intersect(vars_needed, names(TEDS_clean))

df <- TEDS_clean |>
  mutate(
    Tondu = as.numeric(Tondu),
    Tondu = factor(Tondu, levels = 1:7, labels = tondu_labels, ordered = TRUE),
    female = as.factor(female),
    DPP = as.factor(DPP),
    Taiwanese = as.factor(Taiwanese),
    Econ_worse = as.factor(Econ_worse)
  ) |>
  select(all_of(analysis_vars))
```

### Frequency tables and descriptive checks

```{r}
# Tondu by key categorical predictors
tabyl(df, Tondu)
if ("female" %in% names(df)) tabyl(df, Tondu, female)
if ("DPP" %in% names(df)) tabyl(df, Tondu, DPP)
if ("Taiwanese" %in% names(df)) tabyl(df, Tondu, Taiwanese)
if ("Econ_worse" %in% names(df)) tabyl(df, Tondu, Econ_worse)
```

```{r}
# Example association test (if both variables exist)
if (all(c("Tondu", "DPP") %in% names(df))) {
  chisq.test(table(df$Tondu, df$DPP))
}
```

### Visualization method

```{r}
ggplot(df, aes(x = Tondu)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Tondu",
    x = "Tondu",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
# Stacked composition plot for DPP by Tondu (if available)
if (all(c("Tondu", "DPP") %in% names(df))) {
  ggplot(df, aes(x = Tondu, fill = DPP)) +
    geom_bar(position = "fill") +
    labs(
      title = "DPP composition within each Tondu category",
      x = "Tondu",
      y = "Proportion"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
}
```

### Modeling method for `Tondu`

Because `Tondu` is ordered, an ordinal logistic regression is appropriate.

```{r}
model_vars <- c("Tondu", "female", "DPP", "age", "income", "edu", "Taiwanese", "Econ_worse")
model_vars <- intersect(model_vars, names(df))

df_tondu <- df |> drop_na(all_of(model_vars))

if (all(c("Tondu", "female", "DPP", "age", "income", "edu", "Taiwanese", "Econ_worse") %in% names(df_tondu))) {
  m_tondu <- MASS::polr(
    Tondu ~ female + DPP + age + income + edu + Taiwanese + Econ_worse,
    data = df_tondu,
    Hess = TRUE
  )
  summary(m_tondu)
}
```

## 7. Next step: analyze `votetsai`

A binary logistic regression is suitable if `votetsai` is a yes/no vote indicator.

```{r}
if ("votetsai" %in% names(df)) {
  df_vote <- df |>
    mutate(votetsai = as.factor(votetsai)) |>
    drop_na(votetsai, Tondu)

  # Crosstab first
  tabyl(df_vote, votetsai, Tondu)

  # Model (adds controls if available)
  vote_predictors <- c("Tondu", "female", "DPP", "age", "income", "edu", "Taiwanese", "Econ_worse")
  vote_predictors <- intersect(vote_predictors, names(df_vote))

  form <- as.formula(paste("votetsai ~", paste(vote_predictors, collapse = " + ")))
  m_vote <- glm(form, data = df_vote |> drop_na(all_of(c("votetsai", vote_predictors))), family = binomial())
  summary(m_vote)
}
```

## 8. Frequency table and barchart of `Tondu` with labels

The provided sample line should use `factor()` for labels (not `as.numeric(..., labels = ...)`).

```{r}
TEDS_2016 <- TEDS_2016 |>
  mutate(
    Tondu = as.numeric(Tondu),
    Tondu = factor(Tondu, levels = 1:7, labels = tondu_labels)
  )
```

```{r}
tabyl(TEDS_2016, Tondu)

ggplot(TEDS_2016, aes(x = Tondu)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Tondu Frequency Bar Chart",
    x = "Tondu",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```
