---
title: "Assignment 3: ML exercise"
format: html
execute:
  warning: false
  message: false
---

## Load the lab and data

The lab script for this assignment is `Lab_ML01.R`. It loads the TEDS 2016 data and walks through exploratory analysis, clustering, and supervised learning models.

```{r}
# Run the full lab script.
lab_file <- if (file.exists("Lab_ML01.R")) {
  "Lab_ML01.R"
} else {
  "assignments/Lab_ML01.R"
}

source(lab_file)
```

For this assignment, I load the TEDS 2016 data directly using the code from the prompt.

```{r}
library(haven)

TEDS_2016 <-
  read_stata("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")
```

## Multiple Regression Model

The dependent variable is `votetsai`, which records whether the respondent voted for Tsai Ing-wen or another candidate. The independent variables are age, education, and income.

```{r}
model_lm <- lm(votetsai ~ age + edu + income, data = TEDS_2016)
summary(model_lm)
```

## Creating a Function: `regplot`

This function combines `lm()`, `plot()`, and `abline()` to create a regression fit plot.

```{r}
regplot <- function(x, y) {
  fit <- lm(y ~ x)
  plot(x, y)
  abline(fit, col = "red")
}
```

## Regression Plots

### Age

```{r}
regplot(TEDS_2016$age, TEDS_2016$votetsai)
```

### Education

```{r}
regplot(TEDS_2016$edu, TEDS_2016$votetsai)
```

### Income

```{r}
regplot(TEDS_2016$income, TEDS_2016$votetsai)
```

## What Is the Problem?

```{r}
table(TEDS_2016$votetsai, useNA = "ifany")
```

The problem is that `votetsai` is a categorical dependent variable with only two categories, usually coded as `0` and `1`. A regular linear regression model assumes that the dependent variable is continuous, but vote choice is binary.

## How Can Prediction Be Improved?

Prediction can be improved by using a model designed for a categorical dependent variable. For a two-category outcome like `votetsai`, logistic regression is more appropriate than linear regression because it predicts probabilities between 0 and 1.

```{r}
model_logit <- glm(
  votetsai ~ age + edu + income,
  data = TEDS_2016,
  family = binomial
)

summary(model_logit)
```

