---
title: "Assignment 4: Text analytics with quanteda"
format: html
execute:
  warning: false
  message: false
---

```{r}
#| label: setup-packages

library(quanteda)
library(quanteda.textstats)
library(quanteda.textplots)
library(quanteda.textmodels)
library(ggplot2)
library(tidyverse)
```

## US presidential inaugural speeches

The `data_corpus_inaugural` object is included with current **quanteda** lib. Older setups used the add-on **quanteda.corpora**; the code below loads whichever is available. After preprocessing, documents are compared with cosine similarity on grouped DFMs and with keyness across historical periods.

```{r}
#| label: inaugural-corpus

corp_inaug <- tryCatch(
  quanteda::data_corpus_inaugural,
  error = function(e) {
    if (!requireNamespace("quanteda.corpora", quietly = TRUE)) {
      stop(
        "Could not load data_corpus_inaugural from quanteda. ",
        "Run install.packages(\"quanteda\") to upgrade, or install.packages(\"quanteda.corpora\").",
        call. = FALSE
      )
    }
    quanteda.corpora::data_corpus_inaugural
  }
)
toks_inaug <- tokens(corp_inaug, remove_punct = TRUE, remove_numbers = TRUE) |>
  tokens_remove(stopwords("english"))
dfmat_inaug <- dfm(toks_inaug)

# One aggregated document per president → similarity between presidencies
dfm_pres <- dfm_group(dfmat_inaug, groups = President)
sim_pres <- textstat_simil(dfm_pres, method = "cosine")
as.matrix(sim_pres)[1:6, 1:6]
```

```{r}
#| label: inaugural-simil-plot

# Cosine similarity as a heatmap-style view (all presidents in grouped DFM)
sim_mat <- as.matrix(sim_pres)
diag(sim_mat) <- NA
ord <- hclust(as.dist(1 - sim_mat), method = "average")$order
sim_ord <- sim_mat[ord, ord]
sim_long <- sim_ord |>
  as.data.frame() |>
  tibble::rownames_to_column("President_A") |>
  pivot_longer(-President_A, names_to = "President_B", values_to = "cosine") |>
  filter(!is.na(cosine))
sim_long$President_A <- factor(sim_long$President_A, levels = rownames(sim_ord))
sim_long$President_B <- factor(sim_long$President_B, levels = colnames(sim_ord))

ggplot(sim_long, aes(x = President_A, y = President_B, fill = cosine)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C", na.value = "grey90") +
  labs(
    title = "Cosine similarity between inaugural corpora (by president)",
    subtitle = "Each cell pools all inaugural speeches for that president; diagonal omitted.",
    x = NULL, y = NULL, fill = "Cosine"
  ) +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Similarity highlights which presidents reuse overlapping vocabulary when pooled across all their inaugurals (length and era affect the measure). Differences overtime can be summmarized by contrasting early versus recent periods.

```{r}
#| label: inaugural-keyness-era

year_doc <- docvars(corp_inaug, "Year")
early <- year_doc < 1900
recent <- year_doc >= 2001
tstat_era <- textstat_keyness(
  dfmat_inaug,
  target = recent,
  reference = early,
  measure = "lr"
)
textplot_keyness(tstat_era, n = 18, margin = 0.15, color = c("steelblue", "tomato3")) +
  labs(
    title = "Keyness: recent inaugurals (2001+) vs. before 1900",
    subtitle = "Which terms are more characteristic of each era (LR)?"
  ) +
  theme_bw(base_size = 12)
```

```{r}
#| label: inaugural-party-keyness

# Party contrasts (Democratic vs Republican inaugurals in this corpus)
party <- docvars(corp_inaug, "Party")
tstat_party <- textstat_keyness(
  dfmat_inaug,
  target = party == "Democratic",
  measure = "lr"
)
textplot_keyness(tstat_party, n = 15, margin = 0.15, color = c("#2171b5", "#cb181d")) +
  labs(title = "Keyness: Democratic vs. Republican inaugural speeches") +
  theme_bw(base_size = 12)
```

**Takeaway.** Inaugurals cluster by president when aggregated, reflecting stable rhetorical habits within a life’s speeches. Keyness separates centuries and parties, surfacing vocabulary that is era-specific (institutions, geopolitics) or aligned with party branding, even though inaugural rhetoric is highly ceremonial.

## Wordfish

## Comparing positions: Wordfish, correspondence analysis, and similarity

-   **Wordfish** gives a **one-dimensional** ideological or emphasis scale with uncertainty and word parameters; good when a single latent dimension dominates (e.g. strategic framing in recurring reports).
-   **Correspondence analysis** (`textmodel_ca()`) finds **low-dimensional** structure (often two axes for plotting) without a Poisson story; axes are orthogonal and interpretable as association patterns, not necessarily a single “left–right” line.
-   **Cosine similarity** (`textstat_simil()`) compares whole **document or group profiles** in feature space but does not produce a latent scale; it is complementary for “who looks like whom.”

## US Department of Defense reports (`wordfish02.R`)

```{r}
#| label: source-wordfish-workshop
source(
  "wordfish02.R",
  encoding = "UTF-8",
  print.eval = TRUE,
  local = knitr::knit_global()
)
```
